---
title: "DiffBind: ChIP-seq pipeline, paired/blocked analysis"
output: 
    html_document:
        toc: true
        toc_float:
           collapsed: false
        number_sections: true
        toc_depth: 3
        fig_width: 7
        fig_height: 6
params:
    csvfile: samplesheet.csv
    contrasts: "group1_vs_group2"
    peakcaller: "macs"
---

<style type="text/css">
   body {
      font-size: 12pt;
   }
</style>

```{r, include=FALSE, warning=FALSE, message=FALSE}
# global variables
dateandtime <- format(Sys.time(), "%a %b %d %Y - %X")
csvfile <- params$csvfile
outbase <- dirname(csvfile)
contrasts <- params$contrasts
peakcaller <- params$peakcaller

# file output suffixes
cp_bed <- "_Diffbind_consensusPeaks_block.bed"
edger_txt <- "_Diffbind_EdgeR_block.txt"
deseq2_txt <- "_Diffbind_Deseq2_block.txt"
edger_bed <- "_Diffbind_EdgeR_block.bed"
deseq2_bed <- "_Diffbind_Deseq2_block.bed"
deseq2_bed_fullist <- "_Diffbind_Deseq2_fullList_block.txt"
edger_bed_fullist <- "_Diffbind_EdgeR_fullList_block.txt"

# knittr configuration
knitr::opts_chunk$set(echo = FALSE, include=TRUE, message=FALSE, warning=FALSE, error=FALSE)

# load libs
suppressMessages(library(DT))
suppressMessages(library(DiffBind))
suppressMessages(library(parallel))
```

**Groups being compared:**
    *`r contrasts`*  
**Peak sources:**
    *`r peakcaller`*  
**Report generated:** 
    *`r dateandtime`*
 
# Peak Data
Read in sample sheet information and peak information
```{r samples} 
samples <- dba(sampleSheet=csvfile)
consensus <- dba.peakset(samples, consensus=DBA_CONDITION)
print(samples)
```

## Correlation heatmap: Only peaks
Pearson correlation of peak positions: all samples versus all samples  
```{r heatmap1}
try(dba.plotHeatmap(samples, main="", cexRow=1, cexCol=1), silent=TRUE)
```

## PCA: Only peaks
Variance of peak positions  
```{r PCA1, fig.height=5,fig.width=5}
try(dba.plotPCA(samples,DBA_CONDITION), silent=TRUE)
```

## Overlapping peak counts
Number of overlapping peaks.  
If the number of samples is greater than 4, a "consensus" peak Venn diagram is created, where
the consensus peak set are the peaks identified in at least 2 samples for that condition. This is different
from the consensus peak set used for differential analyses.
```{r Venn, fig_height=4}
if (nrow(samples$samples) < 5) {
   dba.plotVenn(samples, 1:nrow(samples$samples)) 
} else {
   dba.plotVenn(consensus, consensus$masks$Consensus, main="Binding Site Overlaps: 'consensus', comparing between groups")
   try(dba.plotVenn(samples,samples$masks[[3]],main="Binding Site Overlaps: samples in Group1"), silent=TRUE)
   try(dba.plotVenn(samples,samples$masks[[4]],main="Binding Site Overlaps: samples in Group2"), silent=TRUE)
}
```

# Consensus peaks and counts
Consensus peaks are peaks found in at least two samples, independent of condition.
FRiP is of consensus peaks and will not match FRiP values calculated outside of this tool.
```{r peaksORsummits}
if ( grepl("narrow", samples$samples$Peaks[1]) ) {
   summits <- TRUE
   print ("Narrow peak calling tool.")
   print ("Differential peaks are 250bp upstream and downstream of the summits.")
} else if ( grepl("broad", samples$samples$Peaks[1]) ) {
   summits <- FALSE
   print ("Broad peak calling tool.")
   print ("Differential peaks are consensus peaks.")
} else {
   summits <- FALSE
   print ("Indeterminate peak calling tool.")
   print ("Differential peaks are consensus peaks.")
}
```

```{r DBcount}
if (summits == TRUE) {
	DBdataCounts <- dba.count(samples, summits=250)
} else {
	DBdataCounts <- dba.count(samples)
}
print(DBdataCounts)
outfile2 <- paste0(contrasts, "-", peakcaller, cp_bed)
consensus2 <- dba.peakset(DBdataCounts, bRetrieve=T)
consensus2$name <- paste0("Peak", 1:length(consensus2))
#rtracklayer::export(consensus2, outfile2)
```

## Correlation heatmap: Peaks and reads
Pearson correlation of library-size normalized counts of consensus peaks: all samples versus all samples
```{r heatmap2}
try(dba.plotHeatmap(DBdataCounts, main="", cexRow=1, cexCol=1),silent=TRUE)
```

## Heatmap: Average signal across each peak
1000 most variable consensus peaks (library-size normalized counts)
```{r heatmap3}
try(dba.plotHeatmap(DBdataCounts, correlations=FALSE, cexRow=1, cexCol=1),silent=TRUE)
```

## PCA: Peaks and reads
Variation of library-size normalized counts of consensus peaks
```{r PCA2, fig.height=5,fig.width=5}
try(dba.plotPCA(DBdataCounts, DBA_CONDITION), silent=TRUE)
```

# Set Up Contrast
Contrast is Group1 - Group2.
```{r contrast}
DBdatacontrast <- dba.contrast(DBdataCounts, minMembers=2, categories = DBA_CONDITION,
							   block=DBA_TREATMENT)
print(DBdatacontrast)
```

# Differential Analysis
This report shows the differential analysis with two tools: Deseq2 and EdgeR. For most
projects, Deseq2 is the optimal tool. Both tools assume that the majority of peaks are
not changing between the two conditions. EdgeR also assumes that there are equal numbers
of peaks on each side of the contrast, so it normalizes the data more than Deseq2. EdgeR
is especially useful when this assumption is true or when there are large differences in
library size across samples. All concentrations are on log2 scale.

```{r analyze}
DBAnalysisDeseq2 <- dba.analyze(DBdatacontrast, method = DBA_DESEQ2)
DBAnalysisEdgeR <- dba.analyze(DBdatacontrast, method = DBA_EDGER)
```

```{r report}
DBReportDeseq2 <- dba.report(DBAnalysisDeseq2, method = DBA_DESEQ2_BLOCK)
DBReportEdgeR <- dba.report(DBAnalysisEdgeR, method = DBA_EDGER_BLOCK)
```

## PCA {.tabset .tabset-fade}
Variance of differential peaks only

### DeSeq2 {-}
```{r PCA3, fig.height=5,fig.width=5}
try(dba.plotPCA(DBAnalysisDeseq2, contrast=1, method= DBA_DESEQ2_BLOCK),silent=TRUE)
```

### EdgeR {-}
```{r PCA4, fig.height=5,fig.width=5}
try(dba.plotPCA(DBAnalysisEdgeR, contrast=1, method = DBA_EDGER_BLOCK),silent=TRUE)
```

## MA plot {.tabset .tabset-fade}
"Log concentration" means average concentration across all samples.
Each dot is a consensus peak.

### DeSeq2 {-}
```{r MA_D}
try(dba.plotMA(DBAnalysisDeseq2, method = DBA_DESEQ2_BLOCK),silent=TRUE)
```

### EdgeR {-}
```{r MA_E}
try(dba.plotMA(DBAnalysisEdgeR, method = DBA_EDGER_BLOCK),silent=TRUE)
```

## Volcano plot {.tabset .tabset-fade}
Each dot is a consensus peak.

### DeSeq2 {-}
```{r Volcano1}
try(dba.plotVolcano(DBAnalysisDeseq2, method = DBA_DESEQ2_BLOCK), silent=TRUE)
```

### EdgeR {-}
```{r Volcano2}
try(dba.plotVolcano(DBAnalysisEdgeR, method = DBA_EDGER_BLOCK), silent=TRUE)
```

## Heatmap: Differential {.tabset .tabset-fade}
1000 most significant differential peaks (Deseq2 or EdgeR normalized)

### DeSeq2 {-}
```{r heatmap4D}
try(dba.plotHeatmap(DBAnalysisDeseq2, contrast=1, method = DBA_DESEQ2_BLOCK,
                    correlations=FALSE, margin=20, cexRow=1, cexCol=1), silent=TRUE)
```

### EdgeR {-}
```{r heatmap4E}
try(dba.plotHeatmap(DBAnalysisEdgeR, contrast=1, method = DBA_EDGER_BLOCK,
                    correlations=FALSE, margin=20, cexRow=1, cexCol=1), silent=TRUE)
```

## Top 500 differentially bound peaks {.tabset .tabset-fade}
### DeSeq2 {-}
```{r Deseq2Report}
outfile <- paste0(contrasts, "-", peakcaller, deseq2_txt)
outfile2 <- paste0(contrasts, "-", peakcaller, deseq2_bed)
DBReportDeseq2$name <- paste0("Peak", 1:length(DBReportDeseq2))

tryDeseqExport <- function(DBReportDeseq2, outfile2) {
   tryCatch(
      {
         rtracklayer::export(DBReportDeseq2, outfile2)
      },
      error = function(cond) {
         print("ERROR: Failed to export DeSeq bed file `rtracklayer::export(DBReportDeseq2, outfile2)`, output blank file")
         write.table(outfile2, file='empty', col.names=FALSE)
      }
   )
}

tryDeseqExport(DBReportDeseq2, file.path(outbase, outfile2))

write.table(DBReportDeseq2, file.path(outbase, outfile), quote=F, sep="\t", row.names=F)
D2i <- length(DBReportDeseq2)
if (D2i == 0) {
   i=1
} else if (D2i > 500) {
   i=500
} else {
   i=D2i
}
try(DT::datatable(data.frame(DBReportDeseq2)[1:i,], rownames=F),silent=TRUE)

report2 <- dba.report(DBAnalysisDeseq2,method = DBA_DESEQ2_BLOCK,
                      th=100,bNormalized=T,bFlip=FALSE,precision=0)

outfile3 <- paste0(contrasts, "-", peakcaller, deseq2_bed_fullist)
write.table(report2, file.path(outbase, outfile3), quote=F, sep="\t", row.names=F)
```

### EdgeR {-}
```{r EdgeRReport}
outfile <- paste0(contrasts, "-", peakcaller, edger_txt)
outfile2 <- paste0(contrasts, "-", peakcaller, edger_bed)
DBReportEdgeR$name <- paste0("Peak", 1:length(DBReportEdgeR))

tryEdgeRExport <- function(edger_report, fout) {
   tryCatch(
      {
         rtracklayer::export(edger_report, fout)
      },
      error = function(cond) {
         print("ERROR: Failed to export EdgeR bed file `rtracklayer::export(edger_report, fout))`, output blank file")
         write.table(fout, file='empty', col.names=FALSE)
      }
   )
}

tryEdgeRExport(DBReportEdgeR, file.path(outbase, outfile2))

write.table(DBReportEdgeR, file.path(outbase, outfile), quote=F, sep="\t", row.names=F)
Ei <- length(DBReportEdgeR)
if (Ei == 0) {
   i=1
} else if (Ei > 500) {
   i=500
} else {
   i=Ei
}
try(DT::datatable(data.frame(DBReportEdgeR)[1:i,], rownames=F), silent=TRUE)

report2 <- dba.report(DBAnalysisEdgeR,method = DBA_EDGER_BLOCK,
                      th=100,bNormalized=T,bFlip=FALSE,precision=0)
outfile3 <- paste0(contrasts, "-", peakcaller, edger_bed_fullist)
write.table(report2, file.path(outbase, outfile3), quote=F, sep="\t", row.names=F)
```

## R tool version information
```{r Info}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>