---
title: "DiffBind: cfChIP-seq QC"
output: 
    html_document:
        toc: true
        toc_float:
           collapsed: false
        number_sections: true
        toc_depth: 3
        fig_width: 7
        fig_height: 6
params:
    csvfile: samplesheet.csv
    contrasts: "group1_vs_group2"
    peakcaller: "macs"
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r, include=FALSE, warning=FALSE, message=FALSE}
## grab args
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")

csvfile <- params$csvfile
contrasts <- params$contrasts
peakcaller <- params$peakcaller
```

**Peak sources:**
    *`r peakcaller`*  
**Report generated:** 
    *`r dateandtime`*  

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE, message=FALSE, warning=FALSE, error=FALSE)
suppressMessages(library(DiffBind))
suppressMessages(library(parallel))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(umap))
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))
```

# Peak Data
Read in sample sheet information and peak information
```{r samples} 
samples <- dba(sampleSheet=csvfile)

# if samples have Condition values
if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
  consensus <- dba.peakset(samples,consensus=DBA_CONDITION, minOverlap = min(table(samples$samples$Condition)))
}
print(samples)
```

## Correlation heatmap: Only peaks
Pearson correlation of peak positions: all samples versus all samples  
```{r heatmap1}
try(dba.plotHeatmap(samples,main="",cexRow=1,cexCol=1),silent=TRUE)
```

## PCA: Only peaks
Variance of peak positions  
```{r PCA1, fig.height=5,fig.width=5}
try(dba.plotPCA(samples),silent=TRUE)
```

## Overlapping peak counts
Number of overlapping peaks.  
If the number of samples is greater than 4, a "consensus" peak Venn diagram is created, where
the consensus peak set are the peaks identified in at least 2 samples for that condition. This is different
from the consensus peak set used for differential analyses.
```{r Venn, fig_height=4}
if (nrow(samples$samples) < 5) {
   dba.plotVenn(samples,1:nrow(samples$samples)) 
} else {
   if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
      dba.plotVenn(consensus,consensus$masks$Consensus,main="Binding Site Overlaps: 'consensus', comparing between groups")
   } else {
      print("Consensus peaks were not called")
   }
}
```

# Consensus peaks and counts
Consensus peaks are peaks found in at least two samples, independent of condition.
FRiP is of consensus peaks and will not match FRiP values calculated outside of this tool.
```{r peaksORsummits}
if ( grepl("narrow",samples$samples$Peaks[1]) ) {
   summits <- TRUE
   print ("Narrow peak calling tool.")
   print ("Differential peaks are 250bp upstream and downstream of the summits.")
} else if ( grepl("broad",samples$samples$Peaks[1]) ) {
  summits <- FALSE
  print ("Broad peak calling tool.")
  print ("Differential peaks are consensus peaks.")
} else {
  summits <- FALSE
  print ("Indeterminate peak calling tool.")
  print ("Differential peaks are consensus peaks.")
}
```

```{r DBcount}

if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
  minOv <- min(table(samples$samples$Condition))
} else {
  minOv <- floor(ncol(samples$class)/3)
}

print(paste0("The minimum number of overlaps is: ", minOv))

if (summits == TRUE) {
  DBdataCounts <- dba.count(samples, summits=250, minOverlap = minOv)
} else {
  DBdataCounts <- dba.count(samples, minOverlap = minOv)
}
print(DBdataCounts)

```

## Correlation heatmap: Peaks and reads
Pearson correlation of library-size normalized counts of consensus peaks: all samples versus all samples
```{r heatmap2}
try(dba.plotHeatmap(DBdataCounts,main="",cexRow=1,cexCol=1),silent=TRUE)
```

## Heatmap: Average signal across each peak
1000 most variable consensus peaks (library-size normalized counts)
```{r heatmap3}
try(dba.plotHeatmap(DBdataCounts,correlations=FALSE,cexRow=1,cexCol=1),silent=TRUE)
```

## PCA: Peaks and reads
Variation of library-size normalized counts of consensus peaks
```{r PCA2, fig.height=5,fig.width=5}
try(dba.plotPCA(DBdataCounts),silent=TRUE)
```

```{r TMM}
vec <- c("seqnames", "start", "end", "width", "strand", samples$samples$SampleID)
consensus2 <- dba.peakset(DBdataCounts, bRetrieve=TRUE) %>% ##extracts TMM-normalized  counts
  as.data.frame() %>% setNames(vec) %>% arrange(start, end) %>% mutate(Peaks = paste0("Peak",1:nrow(.))) %>% 
  dplyr::select(1:4, Peaks, samples$samples$SampleID)

outfile1 <- paste0(contrasts, "-", peakcaller, "_DiffBindQC_TMMcounts.csv")
write.csv(consensus2, outfile1, row.names = F)

outfile2 <- paste0(contrasts, "-", peakcaller, "_DiffBindQC_TMMcounts.bed")
write.table(consensus2[,c("seqnames","start","end","Peaks")],
    outfile2, quote=F, sep="\t", row.names=F, col.names=F)

counts_TMM_ALL <- consensus2
rownames(counts_TMM_ALL) <- counts_TMM_ALL$Peaks
counts_TMM_ALL$Peaks <- NULL

counts_TMM_ALL <- counts_TMM_ALL %>% dplyr::select(5:ncol(.)) %>%
  t() %>% log10() %>% as.data.frame(.)
##UMAP coordinates
set.seed(123)
if (nrow(samples$samples) < 16) {
  umap_coord <- umap(counts_TMM_ALL, n_neighbors= nrow(samples$samples)-1)
} else {
  umap_coord <- umap(counts_TMM_ALL)
}
umap_coord <-as.data.frame(umap_coord$layout) %>% setNames(c("UMAP1", "UMAP2"))

outfile <- paste0(contrasts, "-", peakcaller, "_DiffBindQC_UMAP.csv")
write.csv(umap_coord, outfile, row.names = F)
```

## UMAP: peaks and reads 
```{r UMAP_plot}
p <- ggplot(umap_coord,aes(x = UMAP1, y = UMAP2, label = samples$samples$SampleID))+ ##With labels
  geom_point(aes(color=samples$samples$Condition), size = 3) +
  theme_bw()+ ggtitle(paste0("log-transformed counts:", "n = ", nrow(umap_coord))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color = "Phenotypes") + theme(text=element_text(size=15))+
  geom_text_repel(point.size = NA, size = 2.5)
q <- ggplot(umap_coord,aes(x = UMAP1, y = UMAP2)) + ##No labels
  geom_point(aes(color=samples$samples$Condition), size = 3) +
  theme_bw()+ ggtitle(paste0("log-transformed counts:", "n = ", nrow(umap_coord))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color ="Phenotypes") +  theme(text=element_text(size=15))
  ##geom_text_repel(point.size = NA, size = 2.5)
p

if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
q
}
```

## R tool version information
```{r Info}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
