---
title: "DiffBind QC Report"
output: 
    html_document:
        toc: true
        toc_float:
           collapsed: false
        number_sections: true
        toc_depth: 3
        fig_width: 7
        fig_height: 6
params:
    csvfile: ""
    counts_bed: ""
    counts_csv: ""
    peakcaller: ""
    threads: 1
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r, include=FALSE, warning=FALSE, message=FALSE}
## grab args
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")
peakcaller <- params$peakcaller
csvfile <- params$csvfile
counts_bed <- params$counts_bed
counts_csv <- params$counts_csv
threads <- as.integer(params$threads)

# Ensure threads is at least 1
if (is.na(threads) || threads < 1) {
    threads <- 1
}
threads <- as.integer(threads)
options(mc.cores = threads)
```

**Peak sources:**
    *`r peakcaller`*  
**Report generated:** 
    *`r dateandtime`*  

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE, message=FALSE, warning=FALSE, error=FALSE)
suppressMessages(library(DiffBind))
suppressMessages(library(parallel))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(umap))
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))
```

```{r preflight, echo=FALSE, warning=FALSE,message=FALSE}
sample_sheet <- read.csv(csvfile, stringsAsFactors = FALSE)
sample_sheet <- sample_sheet %>%
  tidyr::drop_na(SampleID, bamReads, Peaks) %>%
  dplyr::filter(SampleID != "", bamReads != "", Peaks != "")

# Write empty placeholder outputs and explain why the report was cut short.
# Used whenever the data cannot support a meaningful QC report, so that the
# Snakemake rule still produces every declared output file.
skip_report <- function(reason) {
  cat(
    "\n### DiffBind QC skipped\n",
    reason,
    "\n\nOutputs were written as empty placeholders.\n"
  )

  empty_counts <- data.frame(
    seqnames = character(),
    start = integer(),
    end = integer(),
    width = integer(),
    Peaks = character()
  )

  write.csv(empty_counts, counts_csv, row.names = FALSE)
  write.table(
    empty_counts[, c("seqnames", "start", "end", "Peaks")],
    counts_bed,
    quote = FALSE,
    sep = "\t",
    row.names = FALSE,
    col.names = FALSE
  )
  knitr::knit_exit()
}

if (nrow(sample_sheet) < 2) {
  skip_report("Fewer than 2 valid samples remain after dropping NaN/empty peak rows.")
}
```

```{r functions, echo=FALSE, warning=FALSE,message=FALSE}
remove_negative_coordinates <- function(dbaOBJ) {
  negCoordIdx <- which(dbaOBJ$merged[,2] < 0)
  dbaOBJ$merged[negCoordIdx,2] <- 1
  dbaOBJ$binding[negCoordIdx,2] <- 1
  for (i in seq_along(dbaOBJ$peaks)) {
      dbaOBJ$peaks[[i]]$Start[negCoordIdx] <- 1
  }
  return(dbaOBJ)
}

remove_negative_coordinates2 <- function(dbaOBJ) {
  # This version is to remove negative coordinates deriving from the peak callers
  negCoordIdx <- unique(which(dbaOBJ$merged[,2] <= 0))
  dbaOBJ$merged[negCoordIdx,2] <- 1
  return(dbaOBJ)
}

PCAbasic <- function(pv, max_comps = 3) {
  # Note: PCA is on a log2 scale when counts are available.
  # Returns a data.frame with PC1..PC<max_comps> per sample, or NULL when a PCA
  # cannot be computed (e.g. no consensus sites). With n samples there are at
  # most n components, so missing components are filled with NA rather than
  # indexing past the end of the PCA result.
  if (is.null(pv$binding) || nrow(pv$binding) == 0) {
    print("PCA skipped: no consensus sites in the binding matrix.")
    return(NULL)
  }

  pv <- tryCatch(DiffBind:::pv.pcmask(pv), error = function(e) {
    print(paste0("PCA skipped: ", conditionMessage(e)))
    NULL
  })
  if (is.null(pv) || is.null(pv$pc)) {
    return(NULL)
  }
  pc <- pv$pc

  vr <- pc$sdev ^ 2
  n_avail <- min(max_comps, length(vr), ncol(pc$x))
  if (n_avail < 1) {
    print("PCA skipped: no principal components available.")
    return(NULL)
  }
  pct <- vr[seq_len(n_avail)] / sum(vr) * 100

  plotData <- as.data.frame(matrix(NA_real_, nrow = nrow(pc$x), ncol = max_comps))
  colnames(plotData) <- paste0("PC", seq_len(max_comps))
  rownames(plotData) <- rownames(pc$x)
  plotData[, seq_len(n_avail)] <- pc$x[, seq_len(n_avail), drop = FALSE]

  for (i in seq_len(n_avail)) {
    print(paste0("PC", i, " %: ", pct[i]))
  }
  if (n_avail < max_comps) {
    print(paste0("Only ", n_avail, " principal component(s) available for ",
                 nrow(pc$x), " sample(s); remaining columns are NA."))
  }
  return(plotData)
}

pca_table <- function(pca_df) {
  if (is.null(pca_df)) {
    return(invisible(NULL))
  }
  DT::datatable(pca_df,
    extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')))
}
```

# Peak Data
Read in sample sheet information and peak information
```{r samples}
samples <- dba(sampleSheet=csvfile)

# dba() keeps only sites present in at least 2 samples (default minOverlap).
# If a sample contributes essentially no peaks, the consensus binding matrix
# is empty and every downstream step (PCA, dba.count, TMM export) fails.
if (is.null(samples$binding) || nrow(samples$binding) == 0) {
  peaks_per_sample <- vapply(samples$peaks, nrow, integer(1))
  names(peaks_per_sample) <- samples$samples$SampleID
  print("Peaks per sample:")
  print(peaks_per_sample)
  skip_report(paste0(
    "No consensus sites: no peak was shared by at least 2 samples. ",
    "Peaks per sample: ",
    paste(names(peaks_per_sample), peaks_per_sample, sep = "=", collapse = ", "),
    "."
  ))
}

samples <- remove_negative_coordinates2(samples)
print(samples)
```

## Correlation heatmap: Only peaks
Pearson correlation of peak positions: all samples versus all samples  
```{r heatmap1}
try(dba.plotHeatmap(samples, main="", cexRow=1, cexCol=1),silent=TRUE)
```

## PCA: Only peaks
Variance of peak positions  
```{r PCA1, fig.height=5,fig.width=5}
try(dba.plotPCA(samples),silent=TRUE)

pca1 <- PCAbasic(samples)
pca_table(pca1)
```

## Overlapping peak counts
Number of overlapping peaks, occupancy analysis. Only if number of samples is less than 5. 
```{r Venn, fig_height=4}
if (nrow(samples$samples) < 5) {
  dba.plotVenn(samples, seq_len(nrow(samples$samples))) 
} 
```

## Consensus peak calls per group
The consensus peaks here are the peaks identified in at least 2 samples for a given condition/group.
```{r consensus_peaks}
consensus <- tryCatch(
  dba.peakset(samples, consensus = DBA_CONDITION),
  error = function(e) {
    print(paste0("Per-group consensus skipped: ", conditionMessage(e)))
    NULL
  }
)

# dba.peakset() only adds a consensus peakset for conditions with >= 2 samples.
# When every condition has a single sample it adds nothing and the Consensus
# mask is NULL; subsetting on that yields an empty object and dba() errors.
consensus_mask <- if (is.null(consensus)) NULL else consensus$masks$Consensus
if (!is.null(consensus_mask) && any(consensus_mask)) {
  consensus2 <- dba(consensus, mask = consensus_mask)
  consensus2
} else if (!is.null(consensus)) {
  print("No per-group consensus peaksets: every condition/group has fewer than 2 samples.")
  print(table(samples$samples$Condition))
}
```

# Consensus peaks and counts
Consensus peaks are unions of peaks across samples. To be defined as a consensus peak that
union must overlap a minumum number of peaks across all samples. The minimum
is either the size of the smallest group, or where group information is unavailable, 1/3 
of all samples.  
FRiP is of consensus peaks and will not match FRiP values calculated outside of this tool.
```{r peaksORsummits}
if ( peakcaller == "macsNarrow" ) {
    summits_arg <- 250
    print ("Ran macsNarrow.")
    print ("Output peaks are 250bp upstream and downstream of the summits.")
} else {
    summits_arg <- FALSE
    print ("Assuming broad peak calling tool.")
    print ("Output peaks are consensus peaks.")
}
```

```{r DBcount}
if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
  minOv <- min(table(samples$samples$Condition))
} else {
  minOv <- floor(ncol(samples$class)/3)
}

if (minOv < 2) { minOv <- 2}

print(paste0("The minimum number of overlaps is: ", minOv))

samples$config$cores <- threads
DBdataCounts <- dba.count(samples, summits=summits_arg, bParallel=(threads > 1))

# remove negative coordinates when summits_arg is not FALSE
if ( summits_arg != FALSE ) {
    DBdataCounts <- remove_negative_coordinates(DBdataCounts)
}

print(DBdataCounts)

DBreadNum <- data.frame( ID=DBdataCounts$class["ID",],
                         Reads=DBdataCounts$class["Reads",])
rownames(DBreadNum) <- seq_len(nrow(DBreadNum))
print(DBreadNum)
```

## Correlation heatmap: Peaks and reads
Pearson correlation of library-size normalized counts of consensus peaks: all samples versus all samples
```{r heatmap2}
try(dba.plotHeatmap(DBdataCounts,main="",cexRow=1,cexCol=1),silent=TRUE)
```

## Heatmap: Average signal across each peak
1000 most variable consensus peaks (library-size normalized counts)  
Peaks here can often be driven by a single sample. This may be different from the 
differential peak calls which will involve patterns across groups of samples.
```{r heatmap3}
try(dba.plotHeatmap(DBdataCounts,correlations=FALSE,cexRow=1,cexCol=1, colScheme="YlGnBu"),silent=TRUE)
```

## PCA: Peaks and reads
Variation of library-size normalized counts of consensus peaks
```{r PCA2, fig.height=5,fig.width=5}
try(dba.plotPCA(DBdataCounts),silent=TRUE)

PCA2 <- PCAbasic(DBdataCounts)
pca_table(PCA2)
```

```{r TMM}
vec <- c("seqnames", "start", "end", "width", "strand", samples$samples$SampleID)
consensus2 <- dba.peakset(DBdataCounts, bRetrieve=TRUE) %>% ##extracts TMM-normalized  counts
  as.data.frame() %>% 
  setNames(vec) %>% 
  arrange(start, end) %>% 
  mutate(Peaks = paste0("Peak", seq_len(nrow(.)))) %>% 
  dplyr::select(1:4, Peaks, samples$samples$SampleID)

write.csv(consensus2, counts_csv, row.names = F)

write.table(consensus2[,c("seqnames","start","end","Peaks")],
    counts_bed, quote=F, sep="\t", row.names=F, col.names=F)

counts_TMM_ALL <- consensus2
rownames(counts_TMM_ALL) <- counts_TMM_ALL$Peaks
counts_TMM_ALL$Peaks <- NULL

counts_TMM_ALL <- counts_TMM_ALL %>% dplyr::select(5:ncol(.)) %>%
  t() %>% log10() %>% as.data.frame(.)
##UMAP coordinates
set.seed(123)

if (nrow(samples$samples) < 16) {
  neighbors=nrow(samples$samples)-1
  if (neighbors > 1) {
    umap_coord <- umap(counts_TMM_ALL, n_neighbors=neighbors)
  } else {
    umap_coord <- umap(counts_TMM_ALL, n_neighbors=2)
  }
} else {
  umap_coord <- umap(counts_TMM_ALL)
}
umap_coord <- as.data.frame(umap_coord$layout) %>% 
              setNames(c("UMAP1", "UMAP2"))

# write.csv(umap_coord, umapfile, row.names = F)

```

## UMAP: peaks and reads 
```{r UMAP_plot}
p <- ggplot(umap_coord,aes(x = UMAP1, y = UMAP2, label = samples$samples$SampleID))+ ##With labels
  geom_point(aes(color=samples$samples$Condition), size = 3) +
  theme_bw()+ ggtitle(paste0("log-transformed counts:", "n = ", nrow(umap_coord))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color = "Phenotypes") + theme(text=element_text(size=15))+
  geom_text_repel(point.size = NA, size = 2.5)
q <- ggplot(umap_coord,aes(x = UMAP1, y = UMAP2)) + ##No labels
  geom_point(aes(color=samples$samples$Condition), size = 3) +
  theme_bw()+ ggtitle(paste0("log-transformed counts:", "n = ", nrow(umap_coord))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color ="Phenotypes") +  theme(text=element_text(size=15))
  ##geom_text_repel(point.size = NA, size = 2.5)
#p

#if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
q
#}

DT::datatable(umap_coord,
  extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')))
```

## R tool version information
```{r Info}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
