---
title: "DiffBind QC Report"
output: 
    html_document:
        toc: true
        toc_float:
           collapsed: false
        number_sections: true
        toc_depth: 3
        fig_width: 7
        fig_height: 6
params:
    csvfile: ""
    counts_bed: ""
    counts_csv: ""
    peakcaller: ""
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r, include=FALSE, warning=FALSE, message=FALSE}
## grab args
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")
peakcaller <- params$peakcaller
csvfile <- params$csvfile
counts_bed <- params$counts_bed
counts_csv <- params$counts_csv
```

**Peak sources:**
    *`r peakcaller`*  
**Report generated:** 
    *`r dateandtime`*  

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE, message=FALSE, warning=FALSE, error=FALSE)
suppressMessages(library(DiffBind))
suppressMessages(library(parallel))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(umap))
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))
```

```{r functions, echo=FALSE, warning=FALSE,message=FALSE}
remove_negative_coordinates <- function(dbaOBJ) {
  negCoordIdx <- which(dbaOBJ$merged[,2] < 0)
  dbaOBJ$merged[negCoordIdx,2] <- 1
  dbaOBJ$binding[negCoordIdx,2] <- 1
  for ( i in 1:length(dbaOBJ$peaks) ) {
      dbaOBJ$peaks[[i]]$Start[negCoordIdx] <- 1
  }
  return(dbaOBJ)
}

PCAbasic <- function(pv) {
  # Note: PCA is on a log2 scale when counts are available.
  pv <- DiffBind:::pv.pcmask(pv)
  pc <- pv$pc

  vr <- rep(0,length(pc$sdev))
  for (i in 1:length(vr)) {
      vr[i] <- pc$sdev[i] ^ 2
  }

  comps=1:3
  c1 <- comps[1]
  c2 <- comps[2]
  c3 <- comps[3]

  c1p <- vr[c1] / sum(vr) * 100
  c2p <- vr[c2] / sum(vr) * 100
  c3p <- vr[c3] / sum(vr) * 100

  plotData <- as.data.frame(pc$x[,c(c1,c2,c3)])
  colnames(plotData) <- c("PC1","PC2","PC3")
  print(paste0("PC1 %: ", c1p))
  print(paste0("PC2 %: ", c2p))
  print(paste0("PC3 %: ", c3p))  
  return(plotData)
}
```

# Peak Data
Read in sample sheet information and peak information
```{r samples} 
samples <- dba(sampleSheet=csvfile)
print(samples)
```

## Correlation heatmap: Only peaks
Pearson correlation of peak positions: all samples versus all samples  
```{r heatmap1}
try(dba.plotHeatmap(samples, main="", cexRow=1, cexCol=1),silent=TRUE)
```

## PCA: Only peaks
Variance of peak positions  
```{r PCA1, fig.height=5,fig.width=5}
try(dba.plotPCA(samples),silent=TRUE)

pca1 <- PCAbasic(samples)

DT::datatable(pca1,
  extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')))

```

## Overlapping peak counts
Number of overlapping peaks, occupancy analysis. Only if number of samples is less than 5. 
```{r Venn, fig_height=4}
if (nrow(samples$samples) < 5) {
   dba.plotVenn(samples, 1:nrow(samples$samples)) 
} 
```

# Consensus peaks and counts
Consensus peaks are unions of peaks across samples. To be defined as a consensus peak that
union must overlap a minumum number of peaks across all samples. The minimum
is either the size of the smallest group, or where group information is unavailable, 1/3 
of all samples.  
FRiP is of consensus peaks and will not match FRiP values calculated outside of this tool.
```{r peaksORsummits}
if ( peakcaller == "macsNarrow" ) {
    summits_arg <- 250
    print ("Ran macsNarrow.")
    print ("Output peaks are 250bp upstream and downstream of the summits.")
} else {
    summits_arg <- FALSE
    print ("Assuming broad peak calling tool.")
    print ("Output peaks are consensus peaks.")
}
```

```{r DBcount}
if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
  minOv <- min(table(samples$samples$Condition))
} else {
  minOv <- floor(ncol(samples$class)/3)
}

if (minOv < 2) { minOv <- 2}

print(paste0("The minimum number of overlaps is: ", minOv))

DBdataCounts <- dba.count(samples, summits=summits_arg, bParallel=T)

# remove negative coordinates when summits_arg is not FALSE
if ( summits_arg != FALSE ) {
    DBdataCounts <- remove_negative_coordinates(DBdataCounts)
}

print(DBdataCounts)

DBreadNum <- data.frame( ID=DBdataCounts$class["ID",],
                         Reads=DBdataCounts$class["Reads",])
rownames(DBreadNum) <- 1:nrow(DBreadNum)
print(DBreadNum)
```

## Correlation heatmap: Peaks and reads
Pearson correlation of library-size normalized counts of consensus peaks: all samples versus all samples
```{r heatmap2}
try(dba.plotHeatmap(DBdataCounts,main="",cexRow=1,cexCol=1),silent=TRUE)
```

## Heatmap: Average signal across each peak
1000 most variable consensus peaks (library-size normalized counts)  
Peaks here can often be driven by a single sample. This may be different from the 
differential peak calls which will involve patterns across groups of samples.
```{r heatmap3}
try(dba.plotHeatmap(DBdataCounts,correlations=FALSE,cexRow=1,cexCol=1, colScheme="YlGnBu"),silent=TRUE)
```

## PCA: Peaks and reads
Variation of library-size normalized counts of consensus peaks
```{r PCA2, fig.height=5,fig.width=5}
try(dba.plotPCA(DBdataCounts),silent=TRUE)

PCA2 <- PCAbasic(DBdataCounts)

DT::datatable(PCA2,
  extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')))
```

```{r TMM}
vec <- c("seqnames", "start", "end", "width", "strand", samples$samples$SampleID)
consensus2 <- dba.peakset(DBdataCounts, bRetrieve=TRUE) %>% ##extracts TMM-normalized  counts
  as.data.frame() %>% 
  setNames(vec) %>% 
  arrange(start, end) %>% 
  mutate(Peaks = paste0("Peak", 1:nrow(.))) %>% 
  dplyr::select(1:4, Peaks, samples$samples$SampleID)

write.csv(consensus2, counts_csv, row.names = F)

write.table(consensus2[,c("seqnames","start","end","Peaks")],
    counts_bed, quote=F, sep="\t", row.names=F, col.names=F)

counts_TMM_ALL <- consensus2
rownames(counts_TMM_ALL) <- counts_TMM_ALL$Peaks
counts_TMM_ALL$Peaks <- NULL

counts_TMM_ALL <- counts_TMM_ALL %>% dplyr::select(5:ncol(.)) %>%
  t() %>% log10() %>% as.data.frame(.)
##UMAP coordinates
set.seed(123)

if (nrow(samples$samples) < 16) {
  neighbors=nrow(samples$samples)-1
  if (neighbors > 1) {
    umap_coord <- umap(counts_TMM_ALL, n_neighbors=neighbors)
  } else {
    umap_coord <- umap(counts_TMM_ALL, n_neighbors=2)
  }
} else {
  umap_coord <- umap(counts_TMM_ALL)
}
umap_coord <- as.data.frame(umap_coord$layout) %>% 
              setNames(c("UMAP1", "UMAP2"))

# write.csv(umap_coord, umapfile, row.names = F)

```

## UMAP: peaks and reads 
```{r UMAP_plot}
p <- ggplot(umap_coord,aes(x = UMAP1, y = UMAP2, label = samples$samples$SampleID))+ ##With labels
  geom_point(aes(color=samples$samples$Condition), size = 3) +
  theme_bw()+ ggtitle(paste0("log-transformed counts:", "n = ", nrow(umap_coord))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color = "Phenotypes") + theme(text=element_text(size=15))+
  geom_text_repel(point.size = NA, size = 2.5)
q <- ggplot(umap_coord,aes(x = UMAP1, y = UMAP2)) + ##No labels
  geom_point(aes(color=samples$samples$Condition), size = 3) +
  theme_bw()+ ggtitle(paste0("log-transformed counts:", "n = ", nrow(umap_coord))) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color ="Phenotypes") +  theme(text=element_text(size=15))
  ##geom_text_repel(point.size = NA, size = 2.5)
#p

#if ( sum(samples$class["Condition",] != "") == ncol(samples$class) ) {
q
#}

DT::datatable(umap_coord,
  extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv')))
```

## R tool version information
```{r Info}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
