---
title: "DiffBind: chrom-seek pipeline"
subtitle: "EdgeR"
output: 
    html_document:
        toc: true
        toc_float:
           collapsed: false
        number_sections: true
        toc_depth: 3
        fig_width: 7
        fig_height: 6
params:
    csvfile: samplesheet.csv
    contrasts: "group1_vs_group2"
    peakcaller: "macsNarrow"
    counts: ""
    up_file: ""
    down_file: ""     
    list_file: ""
---

<style type="text/css">
    body {
        font-size: 12pt;
    }
</style>

```{r, include=FALSE, warning=FALSE, message=FALSE}
## grab args
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")

csvfile <- params$csvfile
contrasts <- params$contrasts
peakcaller <- params$peakcaller
peak_counts <- params$counts
up_file <- params$up_file
down_file <- params$down_file
list_file <- params$list_file
```

**Groups being compared:**
    *`r contrasts`*  
**Peak sources:**
    *`r peakcaller`*  
**Report generated:** 
    *`r dateandtime`*  

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE, message=FALSE, warning=FALSE, error=FALSE)
suppressMessages(library(DT))
suppressMessages(library(DiffBind))
suppressMessages(library(parallel))
```

# Peak Data
Read in sample sheet information and peak information
```{r samples}
samples <- dba(sampleSheet=csvfile)
print(samples)
```

## Correlation heatmap: Only peaks
Pearson correlation of peak positions: all samples versus all samples  
```{r heatmap1}
dba.plotHeatmap(samples, main="", cexRow=1, cexCol=1)
```

## PCA: Only peaks
Variance of peak positions  
```{r PCA1, fig.height=5,fig.width=5}
dba.plotPCA(samples, DBA_CONDITION)
```

## Overlapping peak counts
Number of overlapping peaks.  
If the number of samples is greater than 4, a "consensus" peak Venn diagram is created, where
the consensus peak set are the peaks identified in at least 2 samples for that condition. This is different
from the consensus peak set used for differential analyses.
```{r Venn, fig_height=4}
if (nrow(samples$samples) < 5) {
   dba.plotVenn(samples, 1:nrow(samples$samples)) 
} else {
   try(consensus <- dba.peakset(samples,consensus=DBA_CONDITION),silent=TRUE)
   if (exists("consensus")) {
   		dba.plotVenn(consensus,consensus$masks$Consensus,main="Binding Site Overlaps: 'consensus', comparing between groups")
   }
   try(dba.plotVenn(samples,samples$masks[[3]],main="Binding Site Overlaps: samples in Group1"),silent=TRUE)
   try(dba.plotVenn(samples,samples$masks[[4]],main="Binding Site Overlaps: samples in Group2"),silent=TRUE)
}
```

# Consensus peaks and counts
Consensus peaks are peaks found in at least two samples, independent of condition.
FRiP is of consensus peaks and will not match FRiP values calculated outside of this tool.
```{r}
if (peakcaller == "macsNarrow") {
    summits <- TRUE
    print ("Ran macsNarrow.")
    print ("Differential peaks are 250bp upstream and downstream of the summits.")
} else {
    summits <- FALSE
    print ("Assuming broad peak calling tool.")
    print ("Differential peaks are consensus peaks.")
}
```

```{r DBcount}
DBdataCounts <- readRDS(file=peak_counts)
print(DBdataCounts)
```

## Correlation heatmap: Peaks and reads
Pearson correlation of library-size normalized counts of consensus peaks: all samples versus all samples
```{r heatmap2}
dba.plotHeatmap(DBdataCounts, main="", cexRow=1, cexCol=1)
```

## Heatmap: Average signal across each peak
1000 most variable consensus peaks (library-size normalized counts)
```{r heatmap3}
try(dba.plotHeatmap(DBdataCounts, correlations=FALSE, cexRow=1, cexCol=1), silent=TRUE)
```

## PCA: Peaks and reads
Variation of library-size normalized counts of consensus peaks
```{r PCA2, fig.height=5, fig.width=5}
dba.plotPCA(DBdataCounts, DBA_CONDITION)
```

# Set Up Contrast
Contrast is Group1 - Group2.
```{r contrast}
DBdatacontrast <- dba.contrast(DBdataCounts, minMembers=2, categories=DBA_CONDITION)
print(DBdatacontrast)
```

# Differential Analysis
All concentrations are on log2 scale.

```{r analyze}
DBAnalysisEdgeR <- dba.analyze(DBdatacontrast, method = DBA_EDGER)
DBReportEdgeR <- dba.report(DBAnalysisEdgeR, method = DBA_EDGER)
```

## PCA
Variance of differential peaks only

```{r PCA3, fig.height=5,fig.width=5}
try(dba.plotPCA(DBAnalysisEdgeR, contrast=1, method=DBA_EDGER),silent=TRUE)
```

## MA plot
"Log concentration" means average concentration across all samples.
Each dot is a consensus peak.

```{r MA_D}
try(dba.plotMA(DBAnalysisEdgeR, method=DBA_EDGER),silent=TRUE)
```

## Volcano plot
Each dot is a consensus peak.

```{r Volcano1}
try(dba.plotVolcano(DBAnalysisEdgeR, method=DBA_EDGER),silent=TRUE)
```


## Heatmap: Differential
1000 most significant differential peaks

```{r heatmap4D}
try(dba.plotHeatmap(DBAnalysisEdgeR, contrast=1, method=DBA_EDGER, correlations=FALSE, margin=20, cexRow=1, cexCol=1),silent=TRUE)
```

## Top 500 or less differentially bound peaks 

```{r EdgeRReport}
UpPeaks <- DBReportEdgeR[which(DBReportEdgeR$Fold > 0)]
if ( isTRUE(UpPeaks) && nrow(UpPeaks) > 0 ) {
   rtracklayer::export(UpPeaks, up_file)
} else {
   file.create(up_file)
   cat("No **UP** regulated peaks found!", file = up_file, append = TRUE)
}

DownPeaks <- DBReportEdgeR[which(DBReportEdgeR$Fold < 0)]
if ( isTRUE(DownPeaks) && nrow(DownPeaks) > 0 ) {
   rtracklayer::export(DownPeaks, down_file)
} else {
   file.create(down_file)
   cat("No **DOWN** regulated peaks found!", file = down_file, append = TRUE)
}

D2i <- length(DBReportEdgeR)
i <- as.integer(min(c(500, as.integer(max(c(D2i, 1))))))
DT::datatable(data.frame(DBReportEdgeR)[1:i,], rownames=F)

report2 <- dba.report(
               DBAnalysisEdgeR,
               method = DBA_EDGER,
               th=100,
               bNormalized=T,
               bFlip=FALSE,
               precision=0,
               bCalled=T
           )
write.table(report2, list_file, quote=F, sep="\t", row.names=F)
```


## R tool version information
```{r Info}
sessionInfo()
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>